---
title: "Data frames and dplyr"
#subtitle: "Lab at the University of Oregon"
author: "Claire Lepault"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
knit: pagedown::chrome_print
---
exclude: true

```{R, setup, include = F}
library(pacman)
p_load(
  broom, tidyverse,
  ggplot2, ggthemes, ggforce, ggridges, cowplot,
  latex2exp, viridis, extrafont, gridExtra,
  kableExtra, snakecase, janitor,
  DT, data.table, dplyr,
  lubridate, knitr, future, furrr,
  estimatr, FNN, parsnip,
  huxtable, here, magrittr
)
# Define colors
red_pink   = "#e64173"
turquoise  = "#20B2AA"
orange     = "#FFA500"
red        = "#fb6107"
blue       = "#3b3b9a"
green      = "#8bb174"
grey_light = "grey70"
grey_mid   = "grey50"
grey_dark  = "grey20"
purple     = "#6A5ACD"
slate      = "#314f4f"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
  }, 
  scipen=999)

options(knitr.table.format = "html",
        knitr_kable_format_args = list(digits = 2, scientific = FALSE))

```
# Prologue

---
name: schedule

# Schedule

## Last time

Getting to know .mono[R]—objects, functions, *etc.*

## Today

Working with data in .mono[R].
- The `data.frame` class
- The `dplyr` package

## References

Most of these slides come from the courses offered by [Ed Rubin](http://edrub.in/) ([@EC607](https://github.com/edrubin/EC607S23)) and [Grant McDermott](https://grantmcdermott.com/) ([@EC607](https://github.com/uo-ec607/lectures)).


---
#To get started

</br>

1. Ensure `tidyverse` is installed: `install.packages('tidyverse')`

--

2. Load package library: `library(tidyverse)`

--

3. Get information on the `storms` dataset : `?storms`

--

4. Display the first 2 rows of the `storms` dataset: `head(storms,2)`

--

5. Display the number of rows and columns of the `storms` dataset: 

--

```{R,print1, echo = T}
sprintf("%i rows x %i columns", nrow(storms), ncol(storms))
```


---
layout: true
# Data frames
---

class: inverse, middle

---

name: df

A `data.frame` is .mono[R]'s base, spreadsheet-like object that holds variables.

--

.ex[Example]

```{R, df_storms}
storms[1:8,c(1:8, 10:11)]
```

---
name: creation

## Creation

The `data.frame()` function creates...
--
 `data.frame` objects.

--

You'll generally define data frames by passing the function<br>(.hi-slate[1]) column names and (.hi-slate[2]) values for the columns.

```{R, ex_creat1, eval = F}
data.frame(var1 = 1:5, var2 = "apple", var3 = rnorm(5))
```

--

You can also assign the values using already-existing objects, _e.g._,

```{R, ex_create2, eval = F}
# An object with value
tmp <- rnorm(5)
# Creating the data frame
data.frame(var1 = 1:5, var2 = "apple", var3 = tmp)
```
---

## Creation

```{R, ex_create3}
# Creating the data frame
our_df <- data.frame(var1 = 1:5, var2 = "apple", var3 = rnorm(5))
# Displaying the dataframe
our_df
```
--
Notice that .mono[R] assumes we want to repeat `"apple"` for the entire column.


---

## Creation

You can also create data frames from other objects (_e.g._, matrices) using the function `as.data.frame()`<sup>.pink[†]</sup>.

.footnote[.pink[†] Or just plain, old `data.frame()`.]

However, your data frame's columns will only have names if your matrix's columns had names.

---
name: names

## Names (of columns)

The columns (variables) in your data frame have names.<sup>.pink[†]</sup>

.footnote[.pink[†] If you don't name the columns, then .mono[R] will.]

.qa[Q] What if you want to see/know those names?

--

.qa[A] You've got a few options.

--

1. The `names()` function returns the *names* of an object. Try `names(our_df)`

--

2. `head(our_df, x)` will show you the first `x` rows of `our_df`.

--

3. In .mono[RStudio]: `View(our_df)` or look in your .mono[Environment] tab.
---

## Naming

The `names()` function will also help you rename any/all variables.

--

Change the names of .b[all variables] (include a name for each variable):
```{R, change_names_all, echo = T}
# Set new names
names(our_df) <- c("name1", "name2", "name3")
names(our_df)
```

--

Change the name of .b[the second variable] (only):
```{R, change_names_one, echo = T}
# Set new names
names(our_df)[2] <- "col2"
names(our_df)
```
---
name: indexing

## Indexing

.purple[Option 1] Index data frames just as you index matrices in .mono[R].
- `our_df[1,1]` grabs the value in the first row of the first variable.
- `our_df[2,]` returns the second row of `our_df` (as a data frame).
- `our_df[,3]` returns the third column (`name3`) of `our_df` (as a vector).

--

.purple[Option 2] Reference values/variables using columns' names.
- `our_df$name1` returns the column named `name1` (as a vector). .hi[Top:] `$`
- `our_df[,"name1"]` returns the column named `name1` (as a vector).
- `our_df["name1"]` returns the column named `x` (as a data frame).
- `our_df[,c("name1","col2")]` returns a data frame with variables `"name1"` and `"col2"`.
---

name: adding

## Adding variables

Just as we referenced .pink[existing] variables using `$var_name`,
<br>we can create .purple[new] varirables using `$new_var`, _e.g._,

```{R, create_var, eval = F}
# Add a variable to our_df
our_df$new_var <- 6:10
```

--

If you want to use existing columns to create a new variable

```{R, create_var2, eval = F}
# Create interaction: multi = var1 * new_var
our_df$multi <- our_df$name1 * our_df$new_var
```
--

.qa[Q] Isn't there a better/faster/less-typing way?

--

.qa[A] Yes. *Enter* `dplyr`
--
 (also: `data.table`, which we'll leave for the future).
 
 
---
layout: true
# dplyr
---

class: inverse, middle
---
name: dplyr

.b.slate[Introduction]

- It's is a package.

--

`dplyr` is part of the [`tidyverse`](https://dplyr.tidyverse.org/) (Hadleyverse), and it follows a grammar-based approach to programming/data work.

--

- `data` compose the subjects of your stories

- `dplyr` provides the *verbs* (action words)
:<br> `filter()`, `mutate()`, `select()`, `group_by()`, `summarize()`, `arrange()`

--

<br>

.hi-slate[*Bonus*] `dplyr` is pretty fast and able to interact with SQL databases.
 
---
layout: true
# Pipes and dplyr
---
class: inverse, middle

---
name: pipes

.b.slate[Introduction]

1. Pipes (`%>%`) make your life easier..super[.pink[†]]

1. `dplyr` is your data-work friend.

.footnote[.pink[†] Check out `magrittr` for more pipe options, _e.g._, `%<>%`.]
---

.b.slate[What is a pipe?]

--

Pipes are a .pink[simplifying] programming tool; make your code easier to read

--

Take the .hi-pink[output] of a function as the .hi-purple[input/argument] of another function

--

In `dplyr`, the expression for a pipe is `%>%`

.footnote[.pink[†] `|>` native pipe as of .mono[R] 4.1.0
]

--

.mono[R]'s pipe specifically plugs the returned object to the .pink[left] of the pipe into the first argument of the function on the .purple[right] fo the pipe, _e.g._,

--

```{R, ex_pipe_simple}
rnorm(10) %>% mean()
```
---

.b.slate[Pipes]

Pipes avoid nested functions, prevent excessive writing to your disc, and increase the readability of our .mono[R] scripts

--

By default, .mono[R] pipes the output from the LHS of the pipe into<br>the .hi[first] argument of the function on the RHS of the pipe.

--

*E.g.*, `a %>% fun(3)` is equivalent to `fun(arg1 = a, arg2 = 3)`.

--

</br>

Think [russian dolls](https://en.wikipedia.org/wiki/Matryoshka_doll)


---

layout: true
# Key dplyr verbs using the pipe

---
class: inverse, middle

---

There are five key dplyr verbs that you need to learn.

1. `select`: Select (i.e. subset) columns by their names. 

2. `filter`: Filter (i.e. subset) rows based on their values.

3. `arrange`: Arrange (i.e. reorder) rows based on their values.

4. `mutate`: Create new columns.

5. `summarize`: Collapse multiple rows into a single summary value.<sup>1</sup>

.footnote[
<sup>1</sup> `summarise` with a "s" works too.
]

--
</br>

Let's practice these commands together using the `storms` data frame that comes pre-packaged with dplyr. 


---
layout: false
name: select

# 1) select

Use commas to select multiple columns out of a data frame. (You can also use "first:last" for consecutive columns). Deselect a column with "-".


```{r select1}
storms %>% 
  select(name:hour, wind, pressure, -day)%>%
  head(8)
```

---

# 1) select *cont.*

The `select(contains(PATTERN))` option provides a nice shortcut in relevant cases.


```{r select2}
storms %>% 
  select(name, contains("diameter"))%>%
  tail(5) #tail shows you the last x (here 5) rows of the dataframe
```


---

# 1) select *cont.*

You can also rename some (or all) of your selected variables in place.
```{r select3}
storms %>%
  select(alias=name, year, month, day, hour,
         wind_speed=wind, ts_diameter=tropicalstorm_force_diameter)%>%
  head(4) 
```

--

If you just want to rename columns without subsetting them, you can use `rename`. Try this now by replacing `select(...)` in the above code chunk with `rename(...)`.


---
name:filter

# 2) filter

We can chain multiple filter commands with the pipe (`%>%`), or just separate them within a single filter command using commas.

.pull-left[
```{r filter1, eval=F}
storms %>% 
  select(name,year,month,day)%>%
  filter(year==2008,
         month >= 6)%>%
  head(3)
```
]

--
.pull-right[

```{r filter1_res, echo=F, results = 'asis'}
storms %>% 
  select(name,year,month,day)%>%
  filter(year==2008,
         month >= 6)%>%
  head(3) %>% 
  DT::datatable(rownames = F, options = list(dom = 't'))
```
]

--

We can also use the `&` ("and") and/or `|` ("or") logical operators.

.pull-left[
```{r filter2, eval=F}
storms %>% 
  select(name,year,month,day)%>%
  filter(year==2008 | 
         month >= 6)%>%
  head(3)
```
]

--

.pull-right[

```{r filter2_res, echo=F, results = 'asis'}
storms %>% 
  select(name,year,month,day)%>%
  filter(year==2008 | 
         month >= 6)%>%
  head(3) %>% 
  DT::datatable(rownames = F, options = list(dom = 't'))
```
]


---

# 2) filter *cont.*

Regular expressions work well too.
```{r filter3}
storms %>% 
  select(name:status)%>%
  filter(grepl("tropical", status))%>%
  head(5) 
```

---

# 2) filter *cont.*

A very common `filter` use case is identifying (or removing) missing data cases. 


```{r filter4}
storms %>% 
  select(name,year,month,day,hour,
         ts_diameter=tropicalstorm_force_diameter)%>%
  filter(is.na(ts_diameter))%>%
  head(1)
```

--
</br>
.hi[To remove missing observations, simply use negation:] `filter(!is.na(ts_diameter))`. Try this yourself.


---
name: arrange

# 3) arrange

```{r arrange1}
storms %>% 
  select(name,year,month,day,hour,
         ts_diameter=tropicalstorm_force_diameter)%>%
  filter(!is.na(ts_diameter))%>%
  arrange(ts_diameter)%>%
  head(3)
```

--

*Note:* Arranging on a character-based column (i.e. strings) will sort alphabetically. Try this yourself by arranging according to the "status" column.

---

# 3) arrange *cont.*

We can also arrange items in descending order using `arrange(desc())`.
```{r arrange2}
storms %>% 
  select(name,year,month,day,hour,
         ts_diameter=tropicalstorm_force_diameter)%>%
  filter(!is.na(ts_diameter))%>%
  arrange(desc(ts_diameter))%>%
  head(3)
```

---
name: mutate

# 4) mutate

You can create new columns from scratch, or (more commonly) as transformations of existing columns.
```{r mutate1}
storms %>% 
  select(name, year, month) %>%
  mutate(decade = paste0(substr(year, start = 1, stop = 3),0),
         quarter = ifelse(month %in% c(12,1,2), "Winter",
                         ifelse(month %in% 3:5, "Spring",
                                ifelse(month %in% 6:8,"Summer","Automn"))),
         text = paste0(name, " occured in the", decade,"'s"))%>%
  head(2)
```
`mutate` is order aware. So you can chain multiple mutates in a single call.

---

# 4) mutate *cont.*

`mutate_if` allows you to easily work on a subset of variables. For example:

```{r, mutate2}
storms %>% 
  select(name:status) %>% 
  mutate(status=as.character(status))%>%
  mutate_if(is.character, toupper) %>%
  head(5)
```

---

# 4) mutate *cont.*

`mutate_at` and `mutate_all` are also helpful. For example:

```{r, mutate3}
storms %>% 
  select(name:status) %>% 
  mutate_at(vars(name,status), list(UP =~ toupper(.))) %>%
  head(3)
```

---
name: summarize

# 5) summarize and group_by()

`summarize()`<sup>.pink[†]</sup> summarizes variables—you choose the variables and the summaries (_e.g._, `mean()` or `min()`). When using `group_by()`, summaries are done by groups:

.footnote[.pink[†] or `summarise()`]

```{r summ1}
storms %>% 
  select(name, year, wind, pressure) %>%
  mutate(decade = paste0(substr(year, start = 1, stop = 3),0))%>%
  group_by(decade) %>% 
  summarize(Nobs = n(),
            mean_wind = mean(wind, na.rm = TRUE),
            max_pressure = max(pressure, na.rm = TRUE))%>%
  head(4)
```

---

# 5) summarize and group_by()

Note that including "na.rm = TRUE" (or,  its alias "na.rm = T") is usually a good idea with summarize functions. Otherwise, any missing value will propogate to the summarized value too.
```{r summ2}
storms %>% 
  summarize(mean_ts_diameter = mean(tropicalstorm_force_diameter))
storms %>% 
  summarize(mean_ts_diameter = mean(tropicalstorm_force_diameter, na.rm = TRUE))
```

---

# 5) summarize and group_by()

The same `if`-based workflow that we saw with `mutate` a few slides back also works with `summarize`. For example:

```{r, summ3}
storms %>% 
  select(name, year, wind, pressure) %>%
  mutate(decade = paste0(substr(year, start = 1, stop = 3),0))%>%
  group_by(decade) %>% 
  summarize_if(is.numeric, mean, na.rm=T) %>%
  head(4)

```

---

# 5) summarize and group_by()

`summarize_at` and `summarize_all` are also useful. For example:

```{r, summ4}
storms %>% 
  select(year, wind, pressure) %>%
  mutate(decade = paste0(substr(year, start = 1, stop = 3),0))%>%
  select(-year)%>%
  group_by(decade) %>% 
  summarize_all(list(Mean=~mean(.,na.rm=TRUE), 
                     Min=~min(.,na.rm=TRUE),
                     Max=~max(.,na.rm=TRUE)))
```

---
name: tidyverse
# The tidyverse

There's more! `dplyr` and .purple[`tidyr`] offer even more....super[.pink[†]]

.footnote[
.pink[†] And these are only two of the packages in the `tidyverse`.
]

- .note[Merging] `inner_join()`,`left_join()`,`anti_join()`
- .note[Viewing data] `glimpse()`, `top_n()`
- .note[Sampling] `sample_n()`, `sample_frac()`
- .note[Summaries] `first()`, `last()`, `nth()`, `n_distinct()`
- .note[Duplicates] `distinct()`
- .note[Missingness] `na_if()`, .purple[`replace_na()`], .purple[`drop_na()`], .purple[`fill()`]
- .note[Tidying]  .purple[`pivot_longer()`], .purple[`pivot_wider()`]


--

The folks at RStudio have put together some great cheatsheets, *e.g.*,
- [`dplyr`](https://raw.githack.com/edrubin/EC524W20/master/resources/cheatsheet-dplyr.pdf)
- [data import](https://raw.githack.com/edrubin/EC524W20/master/resources/cheatsheet-data-import.pdf)
- [data wrangling](https://raw.githack.com/edrubin/EC524W20/master/resources/cheatsheet-data-wrangling.pdf)

---
layout: false
# Table of contents

.col-left[
.small[
#### Prologue
- [To get started](#schedule)

#### Data frames
- [General](#df)
- [Creation](#creation)
- [Naming](#names)
- [Indexing](#indexing)
- [Adding](#adding)
]]

.col-right[
.small[
#### `dplyr`
- [General](#dplyr)
- [Pipes](#pipes)
- [`select()`](#select)
- [`filter()`](#filter)
- [`arrange`](#arrange)
- [`mutate`](#mutate)
- [`summarize()` and `group_by()`](#summarize)
- [The `tidyverse`](#tidyverse)
]
]

